Optimal Symmetric Multimodal Templates and Concatenated Random Forests for Supervised Brain Tumor Segmentation (made easy) with \textit{ANTsR}

### 2014-02-07 paper.tex
ENH:  Major revisions.

\title{Multivariate Neuroimage Analysis with \textit{ANTsR}: Application to Supervised Brain Tumor Segmentation using Random Forests}

\author[label1]{Nicholas J.~Tustison
  \fnref{label0}}
  \fntext[label0]{\scriptsize Corresponding author:  PO Box 801339, Charlottesville, VA 22908; T:  434-924-7730; email address:  ntustison@virginia.edu.
  }
\author[label2]{K.~L.~Shrinidhi}
\author[label1]{Max Wintermark}
\author[label1]{Christopher R.~Durst}
\author[label2]{Jeffrey T.~Duda}
\author[label1]{James C.~Gee}
\author[label3]{Murray C.~Grossman}
\author[label2]{Brian B.~Avants}
\address[label1]{Department of Radiology and Medical Imaging, University of Virginia, Charlottesville, VA}
\address[label2]{Penn Image Computing and Science Laboratory, 
                 Department of Radiology, University of Pennsylvania,
                Philadelphia, PA}
\address[label3]{Department of Neurology, University of Pennsylvania,
                Philadelphia, PA}

@@

\begin{abstract} 
Improvements in image acquisition and increasingly sophisticated statistical techniques 
for medical image analysis have motivated the development of advanced computational software 
packages, many of which
have been made publicly available.  One such popular toolkit is the
free and open source Advanced
Normalization Tools (ANTs) which contains a suite of well-vetted processing tools for
core data transformation tasks such as image registration, image segmentation, 
inhomogeneity field correction, and template building.  Despite the numerous
solutions afforded by medical image analysis software, neuroscience (and data analysis in general) requires
robust, statistical tools for making predictive inferences with respect to given hypotheses
and for rendering quantitative information in clarifying and insightful ways.  
In this paper, we describe the
coupling of ANTs with the widely-used \textit{R} language---an open source environment for 
statistical processing and data visualization---which we denote as \textit{ANTsR}.  One of the
many benefits from such integration is the accessibility to advanced statistical 
and machine learning techniques as well as state-of-the-art visualization possibilities.  
To showcase the flexibility and utility of \textit{ANTsR}, we 
apply it to the difficult problem of brain tumor segmentation from multi-modality
MRI.  The reported algorithmic framework was the top-performing entry in the MICCAI 2013 Multimodal 
Brain Tumor Segmentation (BRATS) challenge and, to our knowledge, is the only entry for
both challenge years (2012, 2013) which has been made publicly available as open source.
\end{abstract}

\begin{keyword}
advanced normalization tools \sep BRATS \sep brain tumor segmentation \sep \textit{R} project \sep random forests
%% keywords here, in the form: keyword \sep keyword
\end{keyword}

@@

\section{Introduction}

%In addition to describing \textit{ANTsR} basics and how it can be
%used generally in multivariate neuroimaging studies, 
We showcase \textit{ANTsR} with
the particularly difficult application of supervised brain tumor
segmentation \citep{bauer2013} for which we use the increasingly popular 
random forests machine learning technique.  Although random forests have
been proposed previously in the literature for, among other things, 
supervised brain segmentation (e.g. \cite{geremia2011}), 
we demonstrate that \textit{ANTsR} significantly facilitates the construction 
of a fully functional, parallelizable workflow for such a task and 
that the performance of our proposed framework has proven extremely effective in 
actual practice, specifically
 in the recent MICCAI 2013 Multimodal Brain Tumor Segmentation
Challenge (see Figure \ref{fig:award}).%
\footnote{
http://martinos.org/qtim/miccai2013/
}
\begin{figure}
  \centering
  \includegraphics[width=90mm]{Figures/award.pdf}
  \caption{First place certificate from the MICCAI 2013 BRATS challenge.}
  \label{fig:award}
\end{figure}

It is interesting to note that our participation in the 2013 challenge
was the result of observing the 2012
challenge, wanting to take advantage of the same methods as the winners,
and finding that no teams had availed their software in any form.  In contrast, 
and very much in line with our philosophy 
regarding open science \citep{tustison2013,ince2012}, we have made this framework available as open 
source (even releasing the code before the actual 2013 contest).%
\footnote{
https://github.com/ntustison/BRATS2013
}
To our knowledge, we are the only competitors out of the 20 or so teams 
that participated in either years' challenges to do this.

@@

@@

Supervised techniques generally consist of a training phase
for model construction followed by prediction using the 
generated model.  For supervised brain tumor segmentation, 
a set of training data consisting of labeled brain images 
(e.g. csf, gray matter, white matter, 
necrosis, edema, enhancing and  non-enhancing tumor) is
used to construct a predictive model.  Although other 
classification techniques have been used to segment
brain tumors (e.g. support vector machines \citep{bauer2011}),
random forest models have proven particularly successful.

@@

\begin{figure}
  \centering
  \begin{tabular}{cc}
    \includegraphics[width=40mm]{Figures/S_templateFA_140.png} &
    \includegraphics[width=40mm]{Figures/S_templateFLAIR_140.png} \\
    (a) & (b) \\
    \includegraphics[width=40mm]{Figures/S_templateT1_140.png} &
    \includegraphics[width=40mm]{Figures/S_templateT2_140.png} \\
    (c) & (d) 
  \end{tabular}
  \caption{Multivariate symmetric template created from the Kirby 
           21 data described in \cite{landman2011}.  Shown are the
           (a) fractional anisotropy (FA), (b) FLAIR, (c) MPRAGE, 
           and (d) T2 template components.
          }
  \label{fig:symmetrictemplates}
\end{figure}

In order to better characterize deviations from normal
multi-modal brain shape and appearance, several features were derived 
using population-specific multivariate template 
construction. A recent neuroimaging reproducibility study
by Landman et al. resulted in an open data cohort of 21
normal individuals, each imaged twice, comprising several
modalities including ASL, FLAIR, DTI, fMRI, T1, and T2 
\cite{landman2011}.  These data (known informally as the
``Kirby'' data set) were selected for deriving
a multivariate template due to its public availability and
inclusion of several modalities (even permitting future 
incorporation into our segmentation framework of modalities 
not currently included with the BRATS challenge).

As detailed in \cite{avants2008,avants2010}, 
given $K$ image modalities for $N$ subjects,  
${\mathbf I} = \{I_1,I_2,\ldots, I_K\}$, multivariate 
template construction iterates between optimizing the set 
of diffeomorphic transforms between the subjects and the 
template, 
$\left\{\left(\phi_1,\phi_1^{-1}\right),\ldots,\left(\phi_N,\phi_N^{-1}\right)\right\}$ 
and constructing the 
optimal multivariate template appearance 
$\mathbf{J}=\{J_1,J_2,\ldots, J_K\}$ to minimize the
following cost function:
\begin{align}
  \sum_{n=1}^N 
        \Bigg[ D &\left( \psi(\mathbf{x}),\phi_1^n(\mathbf{x},1)\right) \\ \nonumber 
        +& \sum_{k=1}^K \lambda_k \Pi_k \left(I_k^n\left(\phi_n(\mathbf{x},0.5)\right),J_k\left(\phi^{-1}_n(\mathbf{x},0.5)\right)\right)\Bigg].
\end{align}
$D$ is the diffeomorphic shape distance,
\begin{align}
D\left( \phi( \mathbf{x},0),\phi( \mathbf{x},1)\right) = \int_0^1 \| \nu(\mathbf{x},t)\|_L dt
\end{align}
dependent on the choice of linear operator, $L$, and $\nu$
is the velocity field
\begin{align}
\nu\left( \phi(\mathbf{x},t) \right) = \frac{d\phi(\mathbf{x},t)}{dt},\,\,\, \phi(\mathbf{x},0) = \mathbf{x}.
\end{align}
Each pairwise registration employing the similarity metric $\Pi_k$ can 
be assigned a relative weighting, $\lambda_k$, to weight a particular
modality's influence in the construction process.

In terms of implementation, this algorithm is 
encapsulated in the script \verb#antsMultivariateTemplateConstruction.sh#,
available in the ANTs repository, which permits parallel processing on
an individual workstation or on a large computational cluster.  In
Figure \ref{fig:symmetrictemplates} we show mid-axial slices from
the Kirby multivariate symmetrical template consisting of FA, FLAIR,  
T1, and T2 components.  Although DTI-based images were not included in the
challenge data, such images have shown discriminative potential 
\citep[e.g.][]{price2003} to warrant investigation in our future
work.  The general procedure for const ructing a symmetric template involves flipping data around the mid-sagittal
plane, as outlined in prior work
\citep{Hopkins2013}. \textcolor{red}{FIXME-nick, is that correct? is
  this covered in the scripts?  if so, where?}

\subsubsection{Multi-Modal Feature Image Generation}

\begin{figure*}
  \centering
  \makebox[\textwidth][c]{
  \includegraphics[width=180mm]{Figures/featureImages.pdf}
  }
  \caption{Representative feature images derived from the 
           BRATS\_HG0301  ``challenge'' data set.
           Neighborhood statistical images for each modality were 
           generated by calculating a given statistic within
           a specified neighborhood radius.  Also calculated for each modality were feature
           images based on either the GMM or the MAP-MRF segmentation.  For the former, we
           show the probability maps for each of the seven labels which are used as feature
           images.  From the resulting hard segmentation, we calculate various geometric 
           measures per connected component of each of the seven labels.  Similarly, the 
           registration to the symmetric template produces the modality-specific 
           difference images with the
           corresponding symmetric template itself and with respect to the 
           contralateral side.   
           This mapping is also used to produce the log Jacobian image.  Finally, the 
           (T1 - TC) image is calculated and, from the
           cerebral mask, we calculate the normalized distance image.  
           }
  \label{fig:featureImages}         
\end{figure*}

For both stages, in contrast to previous generative
modeling approaches for multi-modal tumor segmentation 
\citep[e.g.][]{prastawa2003}, we do not use multivariate 
Gaussians to specify tissue probabilities but rather incorporate each
univariate probability map into the feature vector of the training
data.  As pointed out in \cite{menze2010}, multivariate modeling
might obscure the distinct biological information provided by each 
modality.  Instead, we let the random forest construction 
process determine the optimal combination of such multivariate
information.  Additionally, maximum posterior labeling from both stages
is used to determine the connected components for each label.  
Geometric features (assigned voxel-wise) include the physical volumes 
of each connected component, the volume to surface area ratio, 
the elongation, and eccentricity.

@@

Immediately apparent from these plots are the importance of certain features.
In general, the features based on the symmetric template are quite
discriminative thus justifying the increased computational 
resources required to generate such features.  For single-threaded processing
on the cluster, creating the feature images for a single subject required 
approximately 2 hours of total processing time of which the template registration 
component took approximately 75\%.  Additionally, the first order 
statistical features also seemed fairly important which is something that
previous work had demonstrated \citep[e.g.][]{bauer2012,zikic2012}.
There are a number of differences, however, between low-grade and high-grade 
importance outcomes which would also seem to justify the creation of separate
glioma class models.

>>>>

%\title{Multivariate Neuroimage Analysis with \textit{ANTsR}: Application to Supervised Brain Tumor Segmentation using Random Forests}
\title{Optimal Symmetric Multimodal Templates and Concatenated Random Forests for Supervised Brain Tumor Segmentation (made easy) with \textit{ANTsR}}

\author[label1]{Nicholas J.~Tustison
  \fnref{label0}}
  \fntext[label0]{\scriptsize Corresponding author:  PO Box 801339, Charlottesville, VA 22908; T:  434-924-7730; email address:  ntustison@virginia.edu.
  }
\author[label2]{K.~L.~Shrinidhi}
\author[label1]{Max Wintermark}
\author[label1]{Christopher R.~Durst}
\author[label1]{James C.~Gee}
\author[label3]{Murray C.~Grossman}
\author[label2]{Brian B.~Avants}
\address[label1]{Department of Radiology and Medical Imaging, University of Virginia, Charlottesville, VA}
\address[label2]{Penn Image Computing and Science Laboratory, 
                 Department of Radiology, University of Pennsylvania,
                Philadelphia, PA}
\address[label3]{Department of Neurology, University of Pennsylvania,
                Philadelphia, PA}

@@

\begin{abstract} 
Automated brain tumor segmentation in MRI is complicated by the lack of prior knowledge 
concerning tumor location, spatial extent, shape, possible displacement of normal 
tissue, and intensity signature, all of which limit 
the applicability of well-established techniques that have demonstrated good performance 
in related tasks such as normal brain tissue segmentation.  Due to this 
limitation, we localize our prior knowledge strategy to each subject through the 
use of subject-specific multimodal asymmetrical features which demonstrate excellent
discriminative properties within the random forest machine learning framework.  
This approach is made possible with symmetric multimodal brain templates which
are optimal in terms of both shape and multivariate intensity.  Further performance improvements are demonstrated by a successive application of random forest models for increasing levels of feature refinement 
 to which we refer as {\it concatenated random forests}.  Although the derived workflow involves many disparate components, the newly developed \textit{ANTsR} open source software package significantly facilitates the proposed methodology.  \textit{ANTsR}, a comprehensive interface of the popular Advanced Normalization Tools (ANTs) repository with the \textit{R} statistical project,
integrates well-performing image analysis tools such as image registration, segmentation, bias correction, and template building with access to advanced statistical and machine learning techniques as well as state-of-the-art visualization possibilities.  
The reported algorithmic framework was the top-performing entry in the MICCAI 2013 Multimodal Brain Tumor Segmentation challenge and, to our knowledge, is the only entry for both challenge years (2012 \& 2013) which has been made publicly available.
\end{abstract}

\begin{keyword}
advanced normalization tools \sep BRATS \sep brain tumor segmentation \sep \textit{R} project 
%% keywords here, in the form: keyword \sep keyword
\end{keyword}

@@

\section{Introduction}
Given the complexity of tumor growth and appearance and the need
for precise volumetric measurements for tumor characterization, 
much research has been invested in computational approaches for 
automated segmentation of tumor regions in MR images.  Several 
approaches have been previously proposed in the literature as
detailed in recent reviews \citep{angelini2007,bauer2013}.  
Any type of commensurable algorithmic evaluation, however, is
extremely problematic due to widely varying performance assessments
described in the corresponding publications, lack of publicly available 
evaluation data, and privately kept algorithmic instantiations.  
In response to these issues, the Multimodal Brain Tumor Segmentation 
(BRATS) challenge was initiated in 2012 (and continued in 2013) under 
the auspices of the Medical Image Computing and Computer Assisted 
Intervention Society in association with their annual international 
conference \citep{menze2014}.

After observing the 2012 challenge to gain insight for our own research 
needs, we found that no teams had availed their software in any 
form which prompted our participation in the challenge which took place
the following year in Nagoya, Japan.  Heavily influenced by the success 
of participants in the 2012 challenge (specifically, the work of 
\cite{bauer2012,geremia2012,zikic2012}), we adopted the increasingly popular 
random forest machine learning framework \citep{breiman2001} which 
permits the inclusion of various features which might prove saliently 
discriminative.  Instrumental to our success in the 2013 challenge 
was the use of features based on multimodal shape and appearance 
asymmetries which were generated via the construction of symmetric, 
multivariate brain templates.  Additional improvements in performance
were due to the use of a two-stage random forest model approach
whereby the output probability images from the application of the first
random forest model seeds the construction of a subset of new feature images
for input into the second stage.  The results of the second random forest model
are then refined using a series of heuristically-derived binary morphological
operations.

In line with our philosophy regarding open science \citep{tustison2013,ince2012}, 
we made the code available as open source%
\footnote{
https://github.com/ntustison/BRATS2013
}
and, to our knowledge, we are the only competitors out of the 20 or so teams 
that participated in either years' challenges to do this.  All code is based
on the well-vetted and open source Insight Toolkit (ITK)%
\footnote{
http://www.itk.org/
} 
of the National Institutes of Health and is available through our \textit{ANTsR} 
package.

In subsequent sections, we describe in greater detail the automated 
supervised brain segmentation pipeline including the asymmetry
analysis and the concatenated random forest construction.  We also
describe the code and data used for the BRATS challenge so that 
the interested reader can reproduce our results.  Finally, we 
illustrate performance with results from the BRATS 2013 challenge.

% which has demonstrated good performance
%in other neuro-related tasks \citep[e.g.,][]{geremia2011}.

%Although random forests have been proposed previously in the 
%literature for, among other things, supervised brain segmentation (e.g. 
%\cite{geremia2011}), 
%we demonstrate that \textit{ANTsR} significantly facilitates the construction 
%of a fully functional, parallelizable workflow for such a task and 
%that the performance of our proposed framework has proven extremely effective in 
%actual practice, specifically
% in the recent MICCAI 2013 Multimodal Brain Tumor Segmentation
%Challenge.%
%\footnote{
%http://martinos.org/qtim/miccai2013/
%}
%\begin{figure}
%  \centering
%  \includegraphics[width=90mm]{Figures/award.pdf}
%  \caption{First place certificate from the MICCAI 2013 BRATS challenge.}
%  \label{fig:award}
%\end{figure}

@@

Certain key elements characterize the supervised brain tumor segmentation
protocol including:
\begin{itemize}
  \item construction of symmetric multivariate templates for multimodal 
        asymmetry-based features,
  \item generation of other image-based features,
  \item training of random forest models, and
  \item prediction using the proposed framework.
\end{itemize}
A description of each of these components is detailed below along.

\subsection{Multimodal Asymmetry Features from Multivariate Symmetric Templates}

\setlength{\tabcolsep}{2pt}

\begin{figure}[htb]
  \centering
  \begin{tabular}{cccc}
  \includegraphics[width=20mm]{Figures/BRATS_HG0001/BRATS_HG0001_FLAIR_slice93.png} &
  \includegraphics[width=20mm]{Figures/BRATS_HG0001/BRATS_HG0001_T1_slice93.png} &
  \includegraphics[width=20mm]{Figures/BRATS_HG0001/BRATS_HG0001_T1C_slice93.png} &
  \includegraphics[width=20mm]{Figures/BRATS_HG0001/BRATS_HG0001_T2_slice93.png} \\
  (a) & (b) &
  (c) & (d) \\
  \end{tabular}
  \caption{Induced bilateral asymmetry due to tumor presence causing
  distortion of the plane of symmetry.  Shown are mid-axial slices of
  one of the BRATS 2013 training data 
  (specifically {\tt BRATS\_HG0001}):  (a) FLAIR, (b) T1, (c) T1C, and
  (d) T2.  }
  \label{fig:asymmetry}
\end{figure}

\setlength{\tabcolsep}{6pt}

In order to better characterize deviations from normal brain shape 
and appearance, several image features were derived using symmetric 
population-specific multivariate templates \citep{avants2008,avants2010}.  
For normal neuroanatomy, the use of spatial prior information 
coupled with image normalization capabilities has proven useful 
in producing improved segmentation results of ``expected'' brain tissue
such as cerebrospinal fluid, gray matter, and white matter 
\citep[e.g.,][]{ashburner1997}.  In contrast, accommodating spatial 
priors to model the presence of a possible tumor and its constituent tissue 
components is difficult. However, since the normal brain 
exhibits a bilaterally symmetric organization, we can 
use the presence of asymmetries to differentiate abnormal 
brain tissue (i.e. tumor).  A similar motivation prompted
the identification of the mid-sagittal plane of symmetry \citep{prima2002}
for feature generation in MS lesion \citep{geremia2011} and 
tumor \citep{geremia2012} identification.  However, this
approach does not take into account the displacement of 
normal tissue due to tumor growth causing the mid-sagittal plane
to deform from its planar structure (cf Figure \ref{fig:asymmetry} (a)).

To take into account these potential asymmetries, we 
require a data set with the same modalities as dictated
by the subject image acquisition protocol.  Although it
would be preferable to build population-specific multivariate
templates from normal data using the same acquisition 
parameters, we substituted publicly available data

A recent neuroimaging reproducibility study
by Landman et al. resulted in an open data cohort of 21
normal individuals, each imaged twice, comprising several
modalities including ASL, FLAIR, DTI, fMRI, T1, and T2 
\citep{landman2011}.  These data (known as the
``MMRR'' data set) were selected for deriving
a multivariate template due to its public availability and
inclusion of several modalities (even permitting future 
incorporation of modalities 
not currently included with the BRATS challenge into our 
segmentation framework).

As detailed in \cite{avants2008,avants2010}, 
given $K$ image modalities for $N$ subjects,  
${\mathbf I} = \{I_1,I_2,\ldots, I_K\}$, multivariate 
template construction iterates between optimizing the set 
of diffeomorphic transforms between the subjects and the 
template, 
$\left\{\left(\phi_1,\phi_1^{-1}\right),\ldots,\left(\phi_N,\phi_N^{-1}\right)\right\}$ 
and constructing the 
optimal multivariate template appearance 
$\mathbf{J}=\{J_1,J_2,\ldots, J_K\}$ to minimize the
following cost function:
\begin{align}
  \sum_{n=1}^N 
        \Bigg[ D &\left( \psi(\mathbf{x}),\phi_1^n(\mathbf{x},1)\right) \\ \nonumber 
        +& \sum_{k=1}^K \lambda_k \Pi_k \left(I_k^n\left(\phi_n(\mathbf{x},0.5)\right),J_k\left(\phi^{-1}_n(\mathbf{x},0.5)\right)\right)\Bigg].
\end{align}
$D$ is the diffeomorphic shape distance,
\begin{align}
D\left( \phi( \mathbf{x},0),\phi( \mathbf{x},1)\right) = \int_0^1 \| \nu(\mathbf{x},t)\|_L dt
\end{align}
dependent on the choice of linear operator, $L$, and $\nu$
is the velocity field
\begin{align}
\nu\left( \phi(\mathbf{x},t) \right) = \frac{d\phi(\mathbf{x},t)}{dt},\,\,\, \phi(\mathbf{x},0) = \mathbf{x}.
\end{align}
Each pairwise registration employing the similarity metric $\Pi_k$ can 
be assigned a relative weighting, $\lambda_k$, to weight a particular
modality's influence in the construction process.  Once the multivariate
template has converged (typically four iterations are used), we symmetrize
the template by flipping each image contralaterally and then running the
multivariate template construction a second time using only the multivariate
template and its symmetrical version.

In terms of implementation, this template building algorithm is 
encapsulated in the script \verb#antsMultivariateTemplateConstruction.sh#,
available in the ANTs repository, which permits parallel processing on
an individual workstation or on a large computational cluster.  In
Figure \ref{fig:symmetrictemplates} we show mid-axial slices from
the MMRR multivariate symmetrical template consisting of FA%
\footnote{
Although DTI-based images were not included in the
challenge data, such images have shown discriminative potential 
\citep[e.g.,][]{price2003} warranting investigation in our future
work.  
}, FLAIR,  
T1, and T2 components.  These images have been made publicly available%
\footnote{
http://figshare.com/articles/Kirby\_multivariate\_template/852989.
All 7 components are stored under the more informal ``Kirby'' moniker.
}
along with other processed images such as corresponding
tissue priors and structural labels.

{\color{red}{Need to describe the actual feature images}}

\begin{figure}[htb]
  \centering
    \includegraphics[width=85mm]{Figures/templateKirby.pdf}
  \caption{Multivariate symmetric template created from the MMRR 
           data described in \cite{landman2011}.  Shown are the
           (a) FA, (b) FLAIR, (c) MPRAGE, 
           and (d) T2 template components.
          }
  \label{fig:symmetrictemplates}
\end{figure}

\begin{figure}[!htb]
  \centering
    \includegraphics[width=85mm]{Figures/asymmetryFeatureImages2.pdf}
  \caption{ $S \underset{b}{\leftrightsquigarrow} \leftrightarrow T$
          }
  \label{fig:asymmetryFeatures}
\end{figure}

\subsection{Multi-Modal Feature Image Generation}

\begin{figure*}[htb]
  \centering
  \makebox[\textwidth][c]{
  \includegraphics[width=180mm]{Figures/featureImages.pdf}
  }
  \caption{Representative feature images derived from the 
           BRATS\_HG0301  ``challenge'' data set.
           Neighborhood statistical images for each modality were 
           generated by calculating a given statistic within
           a specified neighborhood radius.  Also calculated for each modality were feature
           images based on either the GMM or the MAP-MRF segmentation.  For the former, we
           show the probability maps for each of the seven labels which are used as feature
           images.  From the resulting hard segmentation, we calculate various geometric 
           measures per connected component of each of the seven labels.  Similarly, the 
           registration to the symmetric template produces the modality-specific 
           difference images with the
           corresponding symmetric template itself and with respect to the 
           contralateral side.   
           This mapping is also used to produce the log Jacobian image.  Finally, the 
           (T1 - TC) image is calculated and, from the
           cerebral mask, we calculate the normalized distance image.  
           }
  \label{fig:featureImages}         
\end{figure*}

For both stages, in contrast to previous generative
modeling approaches for multi-modal tumor segmentation 
\citep[e.g.,][]{prastawa2003}, we do not use multivariate 
Gaussians to specify tissue probabilities but rather incorporate each
univariate probability map into the feature vector of the training
data.  As pointed out in \cite{menze2010}, multivariate modeling
might obscure the distinct biological information provided by each 
modality.  Instead, we let the random forest construction 
process determine the optimal combination of such multivariate
information.  Additionally, maximum posterior labeling from both stages
is used to determine the connected components for each label.  
Geometric features (assigned voxel-wise) include the physical volumes 
of each connected component, the volume to surface area ratio, 
the elongation, and eccentricity.

\subsection{Supervised Brain Tumor Segmentation with Concatenated Random Forests}
\subsection{\textit{ANTsR}:  An ANTs/\textit{R} Interface}
\subsection{MICCAI BRATS 2013 Brain Tumor Data}

\clearpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Supervised techniques generally consist of a training phase
for model construction followed by prediction using the 
generated model.  For supervised brain tumor segmentation, 
a set of training data consisting of labeled brain images 
(e.g., csf, gray matter, white matter, 
necrosis, edema, enhancing and  non-enhancing tumor) is
used to construct a predictive model.  Although other 
classification techniques have been used to segment
brain tumors (e.g., support vector machines \citep{bauer2011}),
random forest models have proven particularly successful.

@@

@@

Immediately apparent from these plots are the importance of certain features.
In general, the features based on the symmetric template are quite
discriminative thus justifying the increased computational 
resources required to generate such features.  For single-threaded processing
on the cluster, creating the feature images for a single subject required 
approximately 2 hours of total processing time of which the template registration 
component took approximately 75\%.  Additionally, the first order 
statistical features also seemed fairly important which is something that
previous work had demonstrated \citep[e.g.,][]{bauer2012,zikic2012}.
There are a number of differences, however, between low-grade and high-grade 
importance outcomes which would also seem to justify the creation of separate
glioma class models.
